Follow-up Samples

### 2017-03-29 submission/manuscript_outline_20161024.tex
incorporate Amanda edits

\textbf{Background:} Colorectal cancer (CRC) is a worldwide health
problem and research suggests a correlation between the fecal bacterial
microbiome and CRC. This study tested the hypothesis that treatment for
adenoma or carcinoma results in changes to the bacterial community.
Specifically, we tried to identify components within the community that
were different before and after treatment of adenoma, advanced adenoma
(Screen Relevant Neoplasia (SRN)), and carcinoma.

\textbf{Results:} There was a larger change in the bacterial community
in response to treatment for carcinoma versus adenoma (P-value
\textless{} 0.05) but not carcinoma versus SRN cases (P-value
\textgreater{} 0.05). But there was a trend for increasingly less
community similarity between initial and follow up samples from adenoma
to SRN to carcinoma. Despite this, no difference was found in the
relative abundance of any OTU before and after treatment for adenoma,
SRN, or carcinoma groups (P-value \textgreater{} 0.05). Using Random
Forest models to assess whether changes in follow up samples were
towards a normal community showed that only those with carcinoma had a
significant decrease in positive probability (P-value \textless{} 0.05);
providing further evidence that treatment has the greatest effect in
those with carcinoma. The adenoma model used a total of 62 OTUs, the SRN
model used a total of 61 OTUs, and the carcinoma model used a total of
59 OTUs. A total of 26 OTUs were common to both models with many
classifying to commensal bacteria (e.g. \emph{Lachnospiraceae},
\emph{Bacteroides}, \emph{Anaerostipes}, \emph{Blautia}, and
\emph{Dorea}). Both chemotherapy and radiation did not provide any
additional changes to the bacterial community in those treated for
carcinoma (Pvalue \textgreater{} 0.05).

\textbf{Conclusions:} Our data partially supports the hypothesis that
the bacterial community changes after treatment. Individuals with
carcinoma had more drastic differences to the overall community then
those with adenoma. Common OTUs to all models were overwhelmingly from
commensal bacteria, suggesting that these bacteria may be important to
initial polyp formation, development of advanced adenoma, and transition
to carcinoma.

@@

Colorectal cancer (CRC) is currently the third most common cause of
cancer deaths {[}1,2{]}. The rate of disease mortality has seen a
significant decrease, thanks mainly to improvements in screening
{[}1{]}. However, despite this improvement there are still approximately
50,000 deaths from the disease per year {[}2{]}.

Recent studies in humans have shown that both the microbiome and
specific members within it correlate with CRC pathogenesis {[}3,4{]}.
Further, bacterial communities have been observed to be altered between
normal and tumor tissue {[}5{]}. Mouse models of CRC have further
demonstrated the importance of the microbiome, both on a community
{[}3,6{]} and species level {[}4{]}, for tumorgenesis. Collectively,
these studies provide a tantalizing link between our gut bacteria and
CRC and suggest that biomarkers using our microbes could be developed.
Indeed, builidng models using 16S rRNA gene sequencing along with
clinical tests such as Fecal Immunoglobulin Test (FIT) result in good
predictions of CRC {[}7,8{]}. Although these studies show how our gut
bacteria can impact CRC progression via a changed community or invasion
by more inflammatory bacteria {[}9{]}. They provide very little
information as to whether these communities change and rebound towards
normal after successful treatment of adenoma, advanced adenoma (Screen
Relevent Neoplasia (SRN)), or carcinoma.

Providing an answer to this question is important because it has far
reaching implications on both how the bacterial community causes the
formation of more polyps {[}3,4{]} and the ability to be able to use the
microbiome as a predictive screening tool {[}7,8{]}. Understanding polyp
formation and transition to SRN and then carinoma is crucial to being
able to understand how to prevent CRC occurence. Response of the
community to treatment is also eqaully important to predictive models
designed for screening purposes since an unresponsive community would
provide little additional information for important events, such as
recurrence {[}10{]}.

Using pre- (initial) and post- (follow up) treatment samples we tested
the hypothesis that treatment causes detectable changes to the
microbiome in those with adenoma, SRN, and carincoma. First, we assessed
differences between initial and follow up samples in adenoma, SRN, or
carinoma using alpha or beta diversity metrics. Second, we explored
whether models built to classify adenoma, SRN, or carcinoma versus
normal were able to identify specific community members that differed
between initial and follow up. We also used these models to assess
whether changes in the community were toward a more normal micorbiome.
Finally, we assessed both whether surgery for adenomas and SRN provided
larger community changes or whether chemotherapy or radiation provided
additive changes to the microbiome over surgerical resection. This study
helps to provide evidence as to whether treatment can influence the
community and if the CRC microbiome, identified in previous studies,
persists after such interventions.

@@

\textbf{\emph{The Bacterial Community:}} Within our 67-person cohort we
tested whether those with adenoma (n = 22), SRN (n = 19), or carcinoma
(n = 26) had any broad differences between their initial and follow up
samples. We found that those with carcinoma had a more dissimilar
bacterial community between their initial and follow up sample than
those with adenoma (P-value \textless{} 0.001) {[}Figure 1A{]}. Although
no significant differences were observed between SRN and carcinoma there
was an increase in the dissimilarity of the initial and follow up
samples from adenoma (0.55 ± 0.21 (mean ± SD)) to SRN (0.65 ± 0.25) to
carcinoma (0.78 ± 0.15) {[}Figure 1A{]}. The bacterial community
structure before and after surgery was visualized using NMDS for adenoma
{[}Figure 1B{]} (PERMANOVA \textgreater{} 0.05), SRN {[}Figure 1C{]}
(PERMANOVA \textgreater{} 0.05), and carcinoma {[}Figure 1D{]}
(PERMANOVA \textless{} 0.05). Interestingly, when initial and follow up
samples were compared, regardless of whether the lesions were adenoma or
carcinoma, there was no significant overall difference in beta diversity
(PERMANOVA \textgreater{} 0.05). There was no difference between initial
and follow up samples when investigating alpha diversity metrics for
adenoma, SRN, or carcinoma for any metric tested {[}Table S1{]}.
Additionally, there was also no difference in the relative abundance of
any OTU between initial and follow up samples for adenoma, SRN, or
carcinoma only {[}Figure S1{]}.

\textbf{\emph{Adenoma Model}} The range of model AUC's from 100 runs of
20 repeated 10 fold cross-validation was 0.62 - 0.72 with the AUC of the
model used for classification havin an AUC of 0.65 {[}Figure S2A{]}.
There was a total of 62 OTUs in this model with the vast majority
classifying to bacteria typically thought of as commensal {[}Figure
S3A{]}. There was a significant difference between the actual and model
predicted group stratification calls (P-value \textless{} 0.05). There
was no significant decrease in the positive probability of adenoma
between initial and follow up samples (P-value \textgreater{} 0.05)
{[}Figure 2A{]}.

\textbf{\emph{SRN Model}} The range of model AUC's from 100 runs of 20
repeated 10 fold cross-validation was 0.68 - 0.77 with the AUC of the
model used for classification having an AUC of 0.73 {[}Figure S2B{]}.
There was a total of 61 OTUs in the SRN model {[}Figure S3B{]}. Similar
to the adenoma model the vast majority of OTUs classified to bacteria
typically thought of as commensal. Also similar to the adenoma model
there was a significant difference between the actual and model
predicted group stratification calls (P-value \textless{} 0.05). There
was no significant decrease in the positive probability of SRN between
initial and follow up samples (P-value \textgreater{} 0.05) {[}Figure
2B{]}.

\textbf{\emph{Carcinoma Model}} The range of model AUC's from 100 runs
of 20 repeated 10 fold cross-validation was 0.84 - 0.9 with the AUC of
the model used for classification being 0.88 {[}Figure S2C{]}.
Interestingly, the AUCs improved from adenoma to SRN to carcinoma
{[}Figure 2{]}. There was a total of 59 OTUs in the carinoma model
{[}Figure S3C{]}. Similar to the adenoma and SRN models the vast
majority of OTUs classified to bacteria typically thought of as
commensal but OTUs that also classified to \emph{Fusobacterium},
\emph{Porphyromonas}, and \emph{Parvimonas} appear to be important for
carcinoma classification {[}Figure S3C{]}. Also similar to the adenoma
and SRN models there was a significant difference between the actual and
model predicted group stratification calls (P-value \textless{} 0.05).
There was a significant decrease in the positive probability of
carcinoma between initial and follow up samples (P-value \textless{}
0.05) {[}Figure 2C{]}; suggesting that the carcinoma samples changed
towards normal after treatment, unlike either adenoma or SRN. The one
indivdiual still positive for carcinoma after treatment had an increase
in carcinoma positive proability on follow up {[}Figure 2C{]}.

\textbf{\emph{Adenoma, SRN, and Carcinoma Common OTUs}} We next wanted
to know what predictors within the adenoma, SRN, and carcinoma models
were similar to each other. The main purpose was to identify which OTUs
could be important at all three stages of disease. When we compared the
three different models with each other there were a total of 26 common
OTUs. Some of the most common taxonomic identifications belonged to
\emph{Bacteroides}, \emph{Blautia}, \emph{Anaerostipes},
\emph{Lachnospiraceae}, and \emph{Dorea}. These along with the vast
majority of the OTUs that were common between these models had
classifications to bacteria typically thought of as commensal {[}Table
S2{]}.

\textbf{\emph{Treatment Affects on Community}} After observing these
these changes from treatment we assessed whether chemotherapy or
radiation, in the carcinoma group, and surgery, in the adenoma group,
impacted the observed results. In the carcinoma group neither
chemotherapy or radiation provided any additive change from initial
sample over surgery alone (P-value \textgreater{} 0.05) {[}Table S3{]}.
For the adenoma group there was a single difference in observed OTUs
(sobs) between those that received surgery and those that did not
(P-value \textless{} 0.05) {[}Table S4{]}. For the surgery comparison,
adenoma and SRN were combined due to the low number of surgery
occurances in these two groups. There was no difference in the
proportion of those receiving surgery between the adenoma and SRN groups
(P-value \textgreater{} 0.05). This data suggests that microbiome
changes observed in the carcinoma group were mostly a result from
surgery and not from chemotherapy or radiation.

@@

This study builds upon previous work from numerous labs that have
considered both how the bacterial community between those with and
without CRC differ and how it might be used as an early screening tool
{[}7,8,11--13{]}. Here we show that the bacterial community changes
towards normal after treatment for carcinoma and that chemotherapy and
radiation did not provide an additive change. Although some of the
important OTUs classified to genera from bacteria considered the usual
suspects (e.g. \emph{Fusobacterium}, \emph{Porphyromonas}, and
\emph{Parvimonas}) many did not. The majority of important OTUs had
taxonomic classifications for resident gut microbes and were common for
the adenoma, SRN, and carcinoma models. This suggests that members
within the commensal community may be the first that change during CRC
pathogenesis. These subtle changes, in turn, could be the first step in
allowing more inflammatory bacteria to gain a foothold within the colon
{[}9{]}.

Unlike previous studies on the microbiome and CRC, ours focuses both on
identifying commonalities and differences within adenoma, SRN, and
carcinoma groups before and after treatment. Although there were
differences for genera associated with specific bacterium linked with
CRC {[}Figure S3{]}. The majority of important OTUs taxonmically
classified to commensal bacteria {[}Figure S3{]}. Although these changes
may be subtle, due to the lack of signifcant difference in the bacterial
community before and after treatment in adenoma and SRN {[}Figure 1{]},
they support the hypothesis that the first members of the community to
change and potentially stay changed even after treatment are those that
are commensal bacteria.

Many of the common OTUs that we identified taxnomically classified to
potential butyrate producers (e.g. \emph{Clostridiales},
\emph{Roseburia}, and \emph{Anaerostipes}) {[}Table S2{]}. Other OTUs
classified to bacteria that are inhibited by polyphenols (e.g.
\emph{Bacteroides}). Both butyrate and polyphenols are thought to be
protective against cancer, in part by reducing inflammation {[}14{]}.
These protective compounds are derived from the breakdown of fiber,
fruits, and vegetables by resident gut microbes. One example of this
potential diet-microbiome-inflammation-polyp axis is that
\emph{Bacteroides}, which was highly prevalent in our models, are known
to be increased in those with high non-meat based protein consumption
{[}15{]}. High protein consumption in general has been linked with an
increased CRC risk {[}16{]}. Conversely, \emph{Bacteroides} are
inhibited by polyphenols which are derived from fruits and vegetables
{[}17{]}. Our data fits with the hypothesis that the microbial
metabolites from breakdown products within our own diet could not only
help to shape the existing community but also have an effect on CRC risk
and disease progression.

A limitation, in our study, was that there was a significant difference
in the time elapsed in the collection of the follow up sample between
adenoma or SRN versus carcinoma (P-value \textless{} 0.05), with time
passed being less for adenoma (255 ± 42 days) and SRN (250 ± 41) than
carcinoma (351 ± 102). These results would indicate that some of the
differences observed between the carcinoma and adenoma groups could be
due to differences in collection time. Specifically, it could confound
the observation that carcinomas changed more than adenomas {[}Figure
1{]}. However, there are two reasons that this may not be the case.
First, the SRN group did not have a significant difference and the
collection time of their follow up sample was less than the adenoma
group. Second, this confounding would not affect the observations where
models were used since they were built using a different cohort
{[}Figures 2 \& S2-S3{]}.

Another limitation was that we do not know whether individuals who were
still classified as positive by the carcinoma model eventually had a
subsequent CRC recurrence. This information would help to strengthen the
case for this model keeping numerous individuals above the cutoff
threshold even though at follow up they were diagnosed as no longer
having carcinoma. This study also drew heavily from those with Caucasian
ancestry making it possible that the observations may not be
representative of those with either Asian or African ancestry. Although
our training and test set are relatively large we still run the risk of
over-fitting or having a model that may not be representative of other
populations. We've done our best to safeguard against this by not only
running 10-fold cross validation but also having over 100 different
80/20 splits to try and mimic the type of variation that might be
expected to occur.

Despite these shortcomings our findings add to the existing scientific
knowledge on CRC and the microbiome: That there is a measurable
difference in the bacterial community after adenoma, SRN or carcinoma
treatment. Further, the ability for machine learning algorithms to take
OTU data and successfully lower positive probability of carcinoma after
treatment provides evidence that there are specific signatures,
attributable to both inflammtory and resident commensal organisms,
associated with treatment. Our data provides evidence that commensal
bacteria may be important in the development of polyps and potentially
the transition from adenoma to carcinoma.

@@

\textbf{\emph{Study Design and Patient Sampling:}} Sampling and design
have been previously reported in Baxter, et al {[}7{]}. Briefly, study
exclusion involved those who had already undergone surgery, radiation,
or chemotherapy, had colorectal cancer before a baseline fecal sample
could be obtained, had IBD, a known hereditary non-polyposis colorectal
cancer, or familial adenomatous polyposis. Samples used to build the
models for prediction were collected either prior to a colonoscopy or
between 1 - 2 weeks after. The bacterial community has been shown to
normalize back to a pre-colonscopy community within this time period
{[}18{]}. Our training cohort consisted of a total of 423 individuals
{[}Table 1{]}. Our study cohort consisted of 67 individuals with an
initial sample as described and a follow up sample obtained between 188
- 546 days after treatment of lesion {[}Table 2{]}. This study was
approved by the University of Michigan Institutional Review Board. All
study participants provided informed consent and the study itself
conformed to the guidelines set out by the Helsinki Declaration.

\textbf{\emph{16S rRNA Gene Sequencing:}} Sequencing was completed as
described by Kozich, et al. {[}19{]}. DNA extraction used the 96-well
Soil DNA isolation kit (MO BIO Laboratories) and an epMotion 5075
automated pipetting system (Eppendorf). The V4 variable region was
amplified and the resulting product was split between three sequencing
runs with normal, adenoma, and carcinoma evenly represented on each run.
Each group was randomly assigned to avoid biases based on sample
collection location. The initial and follow up samples were sequenced on
the same run.

\textbf{\emph{Sequence Processing:}} The mothur software package
(v1.37.5) was used to process the 16S rRNA gene sequences and has been
previously described {[}19{]}. The general workflow using mothur was:
Paired-end reads were first merged into contigs, quality filtered,
aligned to the SILVA database, screened for chimeras, classified with a
naive Bayesian classifier using the Ribosomal Database Project (RDP),
and clustered into Operational Taxonomic Units (OTUs) using a 97\%
similarity cutoff with an average neighbor clustering algorithm. The
number of sequences for each sample was rarefied to 10523 to minimize
uneven sampling.

\textbf{\emph{Lesion Model Creation:}} The Random Forest {[}20{]}
algorithm was used to create the model used for prediction of lesion
(adenoma or carcinoma) with the main training and testing of the model
completed on an independent data set of 423 individuals. This model was
then applied to our 67-person cohort. It should be noted that all
individuals with an adenoma or carcinoma were grouped together to form
the lesion group and the model was not created to find differences
between normal, adenoma, and carcinoma but rather differences between
both adenoma and carcinoma versus normal.

>>>>

\textbf{Background:} Colorectal cancer (CRC) is a worldwide health
problem and research suggests a correlation between the fecal bacterial
microbiome and CRC. Despite this, very little is known about what
happens to the microbiome after treatment for an adenoma or carcinoma.
This study tested the hypothesis that treatment for adenoma or carcinoma
results in changes towards a normal bacterial community. Specifically,
we tried to identify components within the community that were different
before and after treatment of adenoma, advanced adenoma, and carcinoma.

\textbf{Results:} There was a larger change in the bacterial community
in response to treatment for carcinoma versus adenoma (P-value
\textless{} 0.05) but not carcinoma versus advanced adenoma (P-value
\textgreater{} 0.05). There was a trend for increasingly less community
similarity, between samples pre- and post-treatment from adenoma to
advanced adenoma to carcinoma. Despite this, no difference was found in
the relative abundance of any specific OTU before and after treatment
for adenoma, advanced adenoma, or carcinoma groups (P-value
\textgreater{} 0.05). Using Random Forest models to assess whether
changes in post-treatment samples were towards a normal community, only
those with carcinoma had a significant decrease in positive probability
(P-value \textless{} 0.05); providing further evidence that treatment
has the greatest effect in those with carcinoma. The adenoma model used
a total of 62 OTUs, the SRN model used a total of 61 OTUs, and the
carcinoma model used a total of 59 OTUs. A total of 26 OTUs were common
to all three models with many classifying to commensal bacteria (e.g.
\emph{Lachnospiraceae}, \emph{Bacteroides}, \emph{Anaerostipes},
\emph{Blautia}, and \emph{Dorea}). Both chemotherapy and radiation did
not provide any additional changes to the bacterial community in those
treated for carcinoma (Pvalue \textgreater{} 0.05).

\textbf{Conclusions:} Our data partially supports the hypothesis that
the microbiome changes after treatment towards a normal community.
Individuals with carcinoma had more drastic differences to the overall
community than those with adenoma. Common OTUs to all models were
overwhelmingly from commensal bacteria, suggesting that these bacteria
may be important in initial polyp formation, development of advanced
adenoma, and transition to carcinoma.

@@

Colorectal cancer (CRC) is currently the third most common cause of
cancer deaths {[}1,2{]}. The rate of disease mortality has seen a
significant decrease, thanks mainly to improvements in screening
{[}1{]}. However, despite this improvement there are still approximately
50,000 deaths from the disease per year {[}2{]}. Some of these deaths
are due to disease recurrence, with current estimates that 20-30\% of
those undergoing resection having a CRC recurrence {[}3{]}. This is
important since approximately 30-50\% of those who have a recurrence
will die of CRC {[}4{]}. To reduce the burden on the healthcare system,
finding ways to better stratify those at highest risk of recurrence is
of great importance.

Recent studies in humans and mouse models have shown that altered
membership and structure of the gut microbiome correlate with CRC
pathogenesis {[}5--7{]}. Further, bacterial communities on the mucosa
are altered between normal and tumor tissue {[}8{]}. Collectively, these
studies provide a tantalizing link between our gut bacteria and CRC and
suggest the possibility of using biomarkers to diagnose disease. Indeed,
models created with microbiome data and clinical tests, such as Fecal
Immunoglobulin Test (FIT), result in good predictions of CRC {[}9,10{]}.
While these studies show how changed microbial communities or invasion
by inflammatory species can impact CRC progression {[}11{]}, they
provide very little information as to whether these communities change
and rebound towards normal after successful treatment of adenoma,
advanced adenoma, or carcinoma.

Providing an answer to this question is important because it has far
reaching implications on both how the bacterial community causes the
formation of more polyps {[}5,6{]} and the ability to use the microbiome
as a predictive screening tool {[}9,10{]}. Understanding polyp formation
and transition to advanced adenoma and then carcinoma is crucial to
being able to understand how to prevent CRC occurrence. Response of the
community to treatment is also equally important to predictive models
designed for screening purposes since an unresponsive community would
provide little additional information for important events, such as
recurrence {[}12{]}.

Using pre- (initial) and post- (follow up) treatment samples we tested
the hypothesis that treatment causes detectable changes to the
microbiome in those with adenoma, advanced adenoma, and carincoma.
First, we assessed differences between samples pre- and post-treatment
in adenoma, advanced adenoma, or carcinoma using alpha or beta diversity
metrics. Second, we explored whether models built to classify adenoma,
advanced adenoma, or carcinoma versus normal were able to identify
specific community members that differed between initial and follow up.
We also used these models to assess whether changes in the community
were toward a more normal microbiome. Finally, we assessed both whether
surgery for adenomas and SRN provided larger community changes or
whether chemotherapy or radiation provided additive changes to the
microbiome over surgical resection. This study aims to provide evidence
as to whether the altered CRC microbiome persists or shifts back towards
normal after such interventions.

@@

\textbf{\emph{The Bacterial Community:}} Within our 67-person cohort we
tested whether the microbiome in patients with adenoma (n = 22),
advanced adenoma (n = 19), or carcinoma (n = 26) had any broad
differences between pre- or post-treatment. We found that carcinoma
patients had a more dissimilar bacterial community between their initial
and follow up sample than those with adenoma (P-value \textless{} 0.001)
{[}Figure 1A{]}. Although no significant differences were observed
between advanced adenoma and carcinoma there was an increase in the
dissimilarity of the samples pre- and post-treatment from adenoma (0.55
± 0.21 (mean ± SD)) to advanced adenoma (0.65 ± 0.25) to carcinoma (0.78
± 0.15) {[}Figure 1A{]}. The bacterial community structure before and
after surgery was visualized using NMDS for adenoma {[}Figure 1B{]}
(PERMANOVA \textgreater{} 0.05), advanced adenoma {[}Figure 1C{]}
(PERMANOVA \textgreater{} 0.05), and carcinoma {[}Figure 1D{]}
(PERMANOVA \textless{} 0.05). Interestingly, when pre- and
post-treatment samples were compared, regardless of whether they were
adenoma or carcinoma, there was no significant overall difference in
beta diversity (PERMANOVA \textgreater{} 0.05). There was no difference
between pre- and post-treatment samples when investigating alpha
diversity metrics for adenoma, advanced adenoma, or carcinoma {[}Table
S1{]}. Additionally, there was also no difference in the relative
abundance of any specific OTU between pre- and post-treatment samples
for adenoma, advanced adenoma, or carcinoma only {[}Figure S1{]}.

\textbf{\emph{Adenoma Model:}} Using normal and adenoma individauls from
a separate cohort we created an adenoma model to help assess whether
those with adenoma in this cohort changed towards normal. The range of
model AUC's from 100 runs of 20 repeated 10-fold cross-validation was
0.62 - 0.72 with the AUC of the model used for classification having an
AUC of 0.65 {[}Figure S2A{]}. There was a total of 62 OTUs in this model
with the vast majority classifying to bacteria typically thought of as
commensal {[}Figure S3A{]}. There was a significant difference between
the actual and predicted disease classification (P-value \textless{}
0.05). There was also no significant decrease in the positive
probability of adenoma between pre- and post-treatment samples (P-value
\textgreater{} 0.05) {[}Figure 2A{]}.

\textbf{\emph{Advanced Adenoma Model:}} Using normal and advanced
adenoma individuals from a separate cohort we created an advanced
adenoma model to help assess whether those with advanced adenoma in this
cohort changed towards normal. The range of model AUC's from 100 runs of
20 repeated 10 fold cross-validation was 0.68 - 0.77 with the AUC of the
model used for classification having an AUC of 0.73 {[}Figure S2B{]}.
There was a total of 61 OTUs in the advanced adenoma model {[}Figure
S3B{]}. Similar to the adenoma model the vast majority of OTUs
classified to bacteria typically thought of as commensal. Also similar
to the adenoma model there was a significant difference between the
actual and predicted disease classification (P-value \textless{} 0.05)
and no significant decrease in the positive probability of advanced
adenoma between pre- and post-treatment samples (P-value \textgreater{}
0.05) {[}Figure 2B{]}.

\textbf{\emph{Carcinoma Model:}} Using normal and carcinoma individuals
from a separate cohort we created a carcinoma model to help assess
whether those with carcinoma in this cohort changed towards normal. The
range of model AUC's from 100 runs of 20 repeated 10 fold
cross-validation was 0.84 - 0.9 with the AUC of the model used for
classification being 0.88 {[}Figure S2C{]}. Interestingly, the AUCs
improved from adenoma to advanced adenoma to carcinoma {[}Figure 2{]}.
There was a total of 59 OTUs in the carcinoma model {[}Figure S3C{]}.
Similar to the adenoma and SRN models the vast majority of OTUs
classified to bacteria typically thought of as commensal but OTUs that
also classified to \emph{Fusobacterium}, \emph{Porphyromonas}, and
\emph{Parvimonas} are also important for carcinoma classification
{[}Figure S3C{]}. Also, like the adenoma and advanced adenoma models
there was a significant difference between the actual and predicted
disease classification (P-value \textless{} 0.05). There was a
significant decrease in the positive probability of carcinoma between
pre- and post-treatment samples (P-value \textless{} 0.05) {[}Figure
2C{]}; suggesting that the carcinoma samples changed towards normal
after treatment, unlike either adenoma or advanced adenoma. The one
individual still positive for carcinoma after treatment had an increase
in carcinoma positive proability between their pre- and post-treatment
sample {[}Figure 2C{]}.

\textbf{\emph{Adenoma, Advanced Adenoma, and Carcinoma Common OTUs:}}
Next, to identify which OTUs are important at all stages of disease, we
identified common predictive OTUs within the adenoma, advanced adenoma,
and carcinoma models. When we compared the three different models with
each other there were a total of 26 common OTUs. Some of the most common
taxonomic identifications belonged to \emph{Bacteroides},
\emph{Blautia}, \emph{Anaerostipes}, \emph{Lachnospiraceae}, and
\emph{Dorea}. The vast majority of the OTUs that were common between
these models had classifications to bacteria typically thought of as
commensal {[}Table S2{]}.

\textbf{\emph{Treatment Affects on Community:}} After observing these
changes from treatment we assessed the possible confounding impact of
chemotherapy or radiation, in the carcinoma group, and surgical
resection, in the adenoma group, on the observed results. In the
carcinoma group neither chemotherapy nor radiation provided any
additional change from initial sample over surgery alone (P-value
\textgreater{} 0.05) {[}Table S3{]}. For the adenoma group there was a
single difference in observed OTUs (sobs) between those that received
surgerical resection and those that had regular polyp removal (P-value
\textless{} 0.05) {[}Table S4{]}. For the surgical resection comparison,
adenoma and advanced adenoma were combined due to the low number of
resection occurances in these two groups. Using a fisher exact test
there was no difference in the proportion of those receiving surgerical
resection between the adenoma and advanced adenoma groups (P-value
\textgreater{} 0.05). This data suggests that the microbiome changes
observed in the carcinoma group were mostly a result from surgerical
resection and not from chemotherapy or radiation.

@@

This study builds upon previous work from numerous labs that have
considered both how the bacterial community between those with and
without CRC differs and how it might be used as an early screening tool
{[}9,10,13--15{]}. Here we show that the microbiome changes towards
normal after treatment for carcinoma and that chemotherapy and radiation
did not provide an additive change. Although some of the important OTUs
classified to genera from bacteria considered the ``usual suspects''"
(e.g. \emph{Fusobacterium}, \emph{Porphyromonas}, and \emph{Parvimonas})
many did not. The majority of important OTUs had taxonomic
classifications for resident gut microbes and were common for the
adenoma, advanced adenoma, and carcinoma models. This suggests that
members within the commensal community may be the first that change
during CRC pathogenesis. These subtle changes, in turn, could be the
first step in allowing more inflammatory bacteria to gain a foothold
within the colon {[}11{]}.

Unlike previous studies on the microbiome and CRC, ours focuses on
comparing the microbiome during recovery from treatment in adenoma,
advanced adenoma, and carcinoma groups. Although there were differences
for genera associated with specific bacterium linked with CRC {[}Figure
S3{]}, the majority of important OTUs taxonmically classified to
commensal bacteria {[}Figure S3{]}. Although these changes may be
subtle, due to the lack of significant differences in the bacterial
community pre- and post-treatment in adenoma and advanced adenoma
{[}Figure 1{]}; they support the hypothesis that the first members of
the community to change and potentially stay changed even after
treatment are those that are commensal bacteria.

Many of the common OTUs that we identified taxonomically classified to
potential butyrate producers (e.g. \emph{Clostridiales},
\emph{Roseburia}, and \emph{Anaerostipes}) {[}Table S2{]}. Other OTUs
classified to bacteria that are inhibited by polyphenols (e.g.
\emph{Bacteroides}). Both butyrate and polyphenols are thought to be
protective against cancer, in part by reducing inflammation {[}16{]}.
These protective compounds are derived from the breakdown of fiber,
fruits, and vegetables by resident gut microbes. One example of this
potential diet-microbiome-inflammation-polyp axis is that
\emph{Bacteroides}, which was highly prevalent in our models, are known
to be increased in those with high non-meat based protein consumption
{[}17{]}. High protein consumption in general has been linked with an
increased CRC risk {[}18{]}. Conversely, \emph{Bacteroides} are
inhibited by polyphenols which are derived from fruits and vegetables
{[}19{]}. Our data fits with the hypothesis that the microbial
metabolites from breakdown products within our own diet could not only
help to shape the existing community but also have an effect on CRC risk
and disease progression. Within this context the commensal community may
be an important modifiable risk factor for monitoring and preventing CRC
recurrence after treatment.

A limitation, in our study, was that there was a significant difference
in the time elapsed in the collection of the follow up sample between
adenoma or advanced adenoma versus carcinoma (P-value \textless{} 0.05),
with time passed being less for adenoma (255 ± 42 days) and advanced
adenoma (250 ± 41) than carcinoma (351 ± 102). These results would
indicate that some of the differences observed between the carcinoma and
adenoma groups could be due to differences in collection time.
Specifically, it could confound the observation that carcinomas changed
more than adenomas {[}Figure 1{]}. However, there are two reasons that
this may not be the case. First, the advanced adenoma group had a higher
dissimilarity than adenoma but lower dissimilarity then carcinoma and
the collection time to their follow up samples was less than the adenoma
group. Second, this confounding would not affect the observations where
models were used since they were built using a different cohort
{[}Figures 2 \& S2-S3{]}.

Another limitation of this study was that it drew heavily from those
with Caucasian ancestry making it possible that the observations may not
be representative of those with either Asian or African ancestry.
Although our training and test set are relatively large we still run the
risk of over-fitting or having a model that may not be representative of
other populations. We have done our best to safeguard against this by
not only running 10-fold cross validation but also having over 100
different 80/20 splits to try and mimic the type of variation that might
be expected to occur.

Building off of our results, an area for future research, is that within
our study we do not know whether individuals who were still classified
as positive by the carcinoma model eventually had a subsequent CRC
recurrence. This information would help to strengthen the case for this
model keeping numerous individuals above the cutoff threshold even
though at follow up they were diagnosed as no longer having carcinoma.
It would also provide additional evidence that the microbiome could be
used as a risk stratification tool in monitoring recurrence risk and
whether different interventions could potentially change this community
and lower the probability of future reccurence.

Despite the stated shortcomings our findings add to the existing
scientific knowledge on CRC and the microbiome: That there is a
measurable difference in the bacterial community after adenoma, advanced
adenoma, or carcinoma treatment. Further, the ability for machine
learning algorithms to take OTU data and successfully lower positive
probability of carcinoma after treatment provides evidence that there
are specific signatures, attributable to both inflammatory and resident
commensal organisms, associated with treatment. Our data provides
evidence that commensal bacteria may be important in the development of
polyps, potential transition of advanced adenoma to carcinoma, and
recovery of the microbiome in CRC following treatment.

@@

\textbf{\emph{Study Design and Patient Sampling:}} Sampling and design
have been previously reported in Baxter, et al {[}9{]}. Briefly, study
exclusion involved those who had already undergone surgery, radiation,
or chemotherapy, had colorectal cancer before a baseline fecal sample
could be obtained, had IBD, a known hereditary non-polyposis colorectal
cancer, or familial adenomatous polyposis. Samples used to build the
models for prediction were collected either prior to a colonoscopy or
between 1 - 2 weeks after. The bacterial community has been shown to
normalize back to a pre-colonoscopy community within this time period
{[}20{]}. Our training cohort consisted of a total of 423 individuals
{[}Table 1{]}. Our study cohort consisted of 67 individuals with an
initial sample as described and a follow up sample obtained between 188
- 546 days after treatment of lesion {[}Table 2{]}. This study was
approved by the University of Michigan Institutional Review Board. All
study participants provided informed consent and the study itself
conformed to the guidelines set out by the Helsinki Declaration.

\textbf{\emph{16S rRNA Gene Sequencing:}} Sequencing was completed as
described by Kozich, et al. {[}21{]}. DNA extraction used the 96-well
Soil DNA isolation kit (MO BIO Laboratories) and an epMotion 5075
automated pipetting system (Eppendorf). The V4 variable region was
amplified and the resulting product was split between three sequencing
runs with normal, adenoma, and carcinoma evenly represented on each run.
Each group was randomly assigned to avoid biases based on sample
collection location. The pre- and post-treatment samples were sequenced
on the same run.

\textbf{\emph{Sequence Processing:}} The mothur software package
(v1.37.5) was used to process the 16S rRNA gene sequences and has been
previously described {[}21{]}. The general workflow using mothur was:
Paired-end reads were first merged into contigs, quality filtered,
aligned to the SILVA database, screened for chimeras, classified with a
naive Bayesian classifier using the Ribosomal Database Project (RDP),
and clustered into Operational Taxonomic Units (OTUs) using a 97\%
similarity cutoff with an average neighbor clustering algorithm. The
number of sequences for each sample was rarefied to 10523 to minimize
uneven sampling.

\textbf{\emph{Model Building:}} The Random Forest {[}22{]} algorithm was
used to create the model used to create the three models used. The
adenoma model classified normal versus adenoma, advanced adenoma was
normal versus advanced adenoma, and carcinoma was normal versus
carcinoma. The toal number of individuals in this data set was 423
individuals. There were a total of 239 individuals in the adenoma model,
262 individuals in the advanced adenoma model, and 266 indivdiuals in
the carcinoma model {[}Table 1{]}. Each model was then applied to our
67-person cohort testing prediction of adenoma intial (adenoma n = 22)
versus adenoma follow up (adenoma n = 0), advanced adenoma initial
(advanced adenoma n = 19) versus advanced adenoma follow up (advanced
adenoma n = 0), carcinoma initial (carinoma n = 26) versus carcinoma
follow up (carcinoma n = 1).

The model included only OTU data obtained from 16S rRNA sequencing.
Non-binary data was checked for near zero variance and OTUs that had
near zero variance were removed. This pre-processing was performed with
the R package caret (v6.0.73). Optimization of the mtry hyper-parameter
involved making 100 different 80/20 (train/test) splits of the data
where normal and adenoma, normal and advanced adenoma, or normal and
carcinoma were represented in the same proportion within both the whole
data set and the 80/20 split. For each of the different splits, 20
repeated 10-fold cross validation was performed on the 80\% component to
optimize the mtry hyper-parameter by maximizing the AUC (Area Under the
Curve of the Receiver Operator Characteristic). The resulting model was
then tested on the hold out data obtained from the 20\% component. All
three models had an optimized mtry of 2.
